Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-03-18 ms/ms.Rnw
still working

\newcommand{\shir}{\citeauthor{shirreff_transmission_2011}}
\newcommand{\champr}{\citeauthor{champredon_hiv_2013}}
\newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=small]{todonotes}

@@

\todo{BMB: add to this description}
In disease modeling, one of the most important parameter is the initial exponential growth rate. As it is one of the few parameters that can be directly observed, many modelers try to match the initial exponential growth rate to correctly simulate the epidemic. Furthermore, evolution of virulence is sensitive to the initial growth rate. Thus, we scale a parameter so that every run has an equal exponential rate of growth rate.

@@

@@

@@

d_littleR_m <- melt(vir_matS,value.name="Mean.VL")
shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
shirreff_df_m$run <- "Shirreff"
df_tot <- rbind(shirreff_df_m,d_littleR_m)
vlab <- "mean virulence ($\\log_{10}$ SPVL)"
tlab <- "time (years)"
g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
    geom_line()+labs(x=tlab,y=vlab)+
    theme(legend.position=c(0.8,0.2))
###
dimnames(vir_matS2) <- list(Time=tvec,
               run=c("$10^{-3}$", "$10^{-4}$", "$10^{-5}$"))
g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL")
###
dimnames(vir_matS3) <- list(Time=tvec,
                            run=c("$2.5$", "$3$", "$3.5$"))
g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL")
limits <- c(2.5,4.7)
breaks <- seq(limits[1],limits[2], by = 0.5)
g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
    scale_colour_discrete(name="run")
g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
    scale_colour_discrete(name="$I(0)$")
g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
        scale_colour_discrete(name="$\\alpha(0)$")
grid.arrange(g1.y, g2.y, g3.y, nrow=1)
@

@@

Figure to-do:
\begin{itemize}
\item fig 1. add labels? tweak size? remove redundant y axes?
\item fig 3. add random-mixing model
\item fig 4. re-order (reverse legend order), figure out how to tweak random density width
\end{itemize}

>>>>

\newcommand{\shir}{\citeauthor{shirreff_transmission_2011}}
\newcommand{\champr}{\citeauthor{champredon_hiv_2013}}
\newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
\newcommand{\rzero}{{\cal R}_0}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=small]{todonotes}

@@

One of the most difficult parts of our model comparison exercise is coming up with sets of parameters that are commensurate against many different model structures. For the most part, our models are too complex to easily derive analytical correspondences among them. Given a numerical criterion, such as $r$ (initial exponential growth rate) or $\rzero$ (intrinsic reproductive number), we can adjust one or more parameters by brute force to ensure that all of the models match according to that criterion. While $\rzero$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons. First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$ (CITE?). Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves (CITE Ma et al.), while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease (CITE Wallinga, Teunis, Lipsitch, etc.) Thus, we scale a parameter so that every run has an equal exponential rate of growth rate.

@@

\todo{mention that baseline parameters are geometric means of ranges}

@@

Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a). Our model produces lower peak virulence ($\approx ?$ vs. $\approx ?$) and equilibrium virulence $\approx ?$ vs. $\approx?$) than Shirreff's, probably because \ldots \todo{describe more; is this because we can't match both R0 and r? Can we match r exactly?}. Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b). Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence.

For the baseline set of parameters, with the initial epidemic spread rate scaled to be the
\todo{just double-checking: Figure 1 shows results from the implicit model, right?}

@@

d_littleR_m <- melt(vir_matS,value.name="Mean.VL")
shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
shirreff_df_m$run <- "Shirreff"
df_tot <- rbind(shirreff_df_m,d_littleR_m)
vlab <- "mean virulence ($\\log_{10}$ SPVL)"
tlab <- "time (years)"
g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
    geom_line()+labs(x=tlab,y=vlab)+
    theme(legend.position=c(0.8,0.2))
###
dimnames(vir_matS2) <- list(Time=tvec,
               run=c("$10^{-3}$", "$10^{-4}$", "$10^{-5}$"))
g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL")
###
dimnames(vir_matS3) <- list(Time=tvec,
                            run=c("$2.5$", "$3$", "$3.5$"))
g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL")
limits <- c(2.5,4.7)
breaks <- seq(limits[1],limits[2], by = 0.5)
afun <- function(label,size=8) {
    annotate(geom="text",label=label,x=-Inf,y=Inf,
             size=size,vjust=1,hjust=0)
}
g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
    scale_colour_discrete(name="run")+afun("a")
g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
    scale_colour_discrete(name="$I(0)$")+afun("b")
g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
        scale_colour_discrete(name="$\\alpha(0)$")+afun("c")
grid.arrange(g1.y, g2.y, g3.y, nrow=1)
@

@@

Figure to-do:
\begin{itemize}
\item fig 1. tweak size? remove redundant y axes/labels?
\item fig 3. add random-mixing model; remove x-axis label; rename factor to make strip labels prettier
\item fig 4. re-order (reverse legend order), figure out how to tweak random density width
\end{itemize}
